---
title: 'Supplemental Materials: Time-varying responses of lake metabolism to light and temperature'
fontsize: 12pt
geometry: margin=1in,letterpaper
documentclass: article
mainfont: "Times New Roman"
mathfont: "Times New Roman"
graphics: true
colorlinks: true
supplemental: true
mathspec: true
author:
   - name: Joseph Phillips
affil:
   - name: Department of Integrative Biology, University of Wisconsin--Madison
output:
    bookdown::pdf_document2:
        fig_caption: yes
        number_sections: yes
        template: template.tex
        latex_engine: xelatex
        toc: yes
        toc_depth: 2
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, cache = FALSE}
# load packages
suppressPackageStartupMessages({
  library(knitr)
  library(bookdown)
  library(tidyverse)  
  library(lubridate)
})

# set directory
knitr::opts_knit$set(root.dir = normalizePath(".."))

# set theme
theme_set(theme_bw() %+replace% 
            theme(panel.grid = element_blank(),
                  strip.background = element_blank(),
                  legend.margin = margin(0,0,0,0),
                  strip.text = element_text(size=10),
                  legend.text = element_text(size=10),
                  axis.text=element_text(size=10, color="black"),
                  axis.title.y=element_text(angle = 90 ,margin=margin(0,15,0,0)),
                  axis.title.x=element_text(margin=margin(15,0,0,0))))
```

\clearpage

# Data preparation

## Light 

```{r echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE}
# import hobo
hobo = list.files("data/hobo_raw/clean_data", full.names = T) %>%
  lapply(read_csv) %>%
  bind_rows()

# weather
weather = read_csv("data/weather_raw/weather_2010_2018.csv")

# aggregate HOBO by hour
hobo_hourly = hobo %>%
  mutate(year = year(date_time),
         yday = yday(date_time),
         hour = hour(date_time)) %>%
  group_by(year, yday, hour) %>%
  summarize(date_time = min(date_time),
            lux = mean(lux, na.rm=T)) %>%
  ungroup()

# aggregate weather by hour
weather_hourly = weather %>%
  mutate(year = year(Date_Time),
         yday = yday(Date_Time),
         hour = hour(Date_Time)) %>%
  group_by(year, yday, hour) %>%
  summarize(date_time = min(Date_Time),
            temp = mean(temp, na.rm=T),
            wdir = mean(wdir, na.rm=T),
            temp = mean(temp, na.rm=T),
            wspeed = mean(wspeed, na.rm=T),
            gust = mean(gust, na.rm=T),
            RADGL = mean(RADGL, na.rm=T)) %>%
  ungroup()

# combine HOBO and weather data (only keeping necessary variables)
# divide lux by 54 to roughly convert to PAR
# multiply RADGL by 4.57 to roughly convert ot PAR
# Thimijan and Heins 1983.
hobo_weather = hobo_hourly %>%
  mutate(par_hobo = lux/54) %>%
  select(year, yday, hour, par_hobo) %>%
  full_join(weather_hourly %>%
              filter(!(year %in% c(2010, 2011, 2014))) %>%
              mutate(par_weath = 4.57*RADGL) %>%
              select(year, yday, hour, wspeed, par_weath))

# calculate correlation
cor_natural = cor.test(hobo_weather$par_hobo, hobo_weather$par_weath)
cor_log = cor.test(log1p(hobo_weather$par_hobo), log1p(hobo_weather$par_weath))
```

Light readings were taken either every 15 or 30 minutes (depending on the year) with a light/temperature logger (HOBO Pendant, Onset Computer Corporation, MA, USA) attached to the top of the buoy deploying the sonde; I averaged these data to an hourly scale for further analysis. HOBO loggers measure visible light intensity in lux, which is closely related to the spectral range (400-700nm) used to calculate photosynthetically active radiation (PAR; measured in mol-photons m-2 s-1) (Thimijan and Heins 1983). I also obtained downwelling solar irradiance from the local whether station (in Watts). Following Thimijan and Heins 1983, visibile light intensity and downwelling irradiance can be scalarly converted to photon flux density based on an approximate standard spectrum for sunlight. Specifically, the visible intensity is divided by 54 and downwelling irradiance is multiplied by 4.57. 

When approximately converted to PAR, visibile light intensity and downwelling irradiance were strongly correlated ($r$=0.79 on natural scale and $r$=0.89 on log-log scale; Fig \ref{fig:hobo-weather}). However, when hourly values are plotted through time it appears that downwelling irradiance has higher peaks than the visible intensity (Fig \ref{fig:light-hour}), which makes sense as the donwelling irradiance is calculated using a wider spectral range that visible intensity. 

```{r hobo-weather, echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, fig.height=5, fig.width=5, fig.cap="Visible light intensity measured by HOBO logger and downwelling irradiance from weather station, scalarly converted to approximate PAR equivalent."}
hobo_weather %>%
  ggplot(aes(par_weath, par_hobo))+
  geom_point(size = 0.75, alpha = 0.3)+
  scale_y_continuous(expression("Visible Light Intensity ("*mu*mol~photons~m^{-2}~s^{-1}*")"), 
                     trans="log1p", 
                     breaks = c(2, 10, 50, 250, 1250))+
  scale_x_continuous(expression("Downwelling Irradiance ("*mu*mol~photons~m^{-2}~s^{-1}*")"), trans="log1p", 
                     breaks = c(2, 10, 50, 250, 1250))+
  coord_equal()
```

```{r light-hour, echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, fig.height=5, fig.width=6, fig.cap="PAR equivalent for visible light intensity and downwelling irradiance plotted by hour for 10 days in 2013."}
hobo_weather %>%
  filter(year == 2013, yday %in% c(170:180)) %>%
  rename(`Visible Light`=par_hobo, `Downwelling Irradiance`=par_weath) %>%
  gather(var, value, `Visible Light`, `Downwelling Irradiance`) %>%
  mutate(time = yday + hour/24) %>%
  ggplot(aes(time, value, color = var))+
  geom_line()+
  scale_color_manual("",values=c("magenta3","black"))+
  scale_y_continuous(expression("PAR ("*mu*mol~photons~m^{-2}~s^{-1}*")"), limits=c(0,5000))+
  scale_x_continuous("Time (days)",breaks=c(171,174,177,180))+
  theme(legend.position = c(0.75,0.90))
```

\clearpage

## Profiles

Profiles for temperature and DO were taken at half-meter intervals across the entire water column depth (3.3m) at the same location as the sonde approximately weekly from 2013-2018. The range of values for temperature (Fig \ref{fig:do-prof}) and DO (Fig \ref{fig:temp-prof}) were used to inform the filtering of the sonde data, as described in the main text.

DO never dropped below 9mg/L and were mostly clustered within 1102 mg/L. These handheld readings do not capture the full range of variation, but given that Mývatn is quite shallow at the sonde location (3.3m), does not show much vertical variation in DO (e.g. Fig \ref{fig:minidot}), and has relatively modest diel variation (e.g. Fig \ref{fig:do}) in the summer due to the extended day length at high latitudes, it seems very unlikely that the values drop substantially below 9mg/L during the summer.

```{r do-prof, echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, fig.height=5, fig.width=6, fig.cap="Histogram of DO profiles"}
# profiles
read_csv("data/myvatn_profile.csv") %>%
  filter(sta %in% c(3, 33)) %>%
  mutate(year = year(sampledate)) %>%
  ggplot(aes(do))+
  facet_wrap(~year)+
  geom_histogram()+
  ylab("Count")+
  xlab(expression(DO~"("*mg~O[2]~L^{-1}*")"))
```

```{r temp-prof, echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, fig.height=5, fig.width=6, fig.cap="Histogram of temperature profiles"}
# profiles
read_csv("data/myvatn_profile.csv") %>%
  filter(sta %in% c(3, 33)) %>%
  mutate(year = year(sampledate)) %>%
  ggplot(aes(wtemp))+
  facet_wrap(~year)+
  geom_histogram()+
  ylab("Count")+
  xlab(expression(Temperature~(degree~C)))
```

\clearpage

## Filtering

The DO and temperature data from the sonde had occasional extreme values that were likely erroneous and so were filtered before analysis. To provide an objective method for removing these values, I elimited all DO and temperature observations that were beyond 2 standard deviations of the mean for a given day. While this removed some values that were likely not erroneous, the majority of the elimiated values were quite extreme and very likely erronenous (Figs \ref{fig:do-filter-all} and \ref{fig:do-filter-subset}). Overall, this procedure resulted in the removal of 5.6% of observations.

```{r echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE}

# import raw sonde data; calculate 1 sd for oxygen and temp
sonde_raw = list.files("data/sonde_raw/clean_data", full.names = T) %>%
  lapply(read_csv) %>%
  bind_rows() %>%
  mutate(year = year(Date_Time),
         yday = yday(Date_Time)) %>%
  filter(!(year == 2017 & Date_Time > "2017-08-19"),
         is.na(year) == F) %>%
  group_by(year) %>%
  filter(yday > min(yday) + 3,
         yday < max(yday) - 3) %>%
  group_by(year, yday) %>%
  filter(LDO > 5, Temp < 18, TurbSC < 200) %>%
  mutate(LDO_mean = mean(LDO, na.rm=T),
         LDO_sd = sd(LDO, na.rm=T),
         Temp_mean = mean(Temp, na.rm=T),
         Temp_sd = sd(Temp, na.rm=T)) %>%
  ungroup() %>%
  mutate(do_filt = ifelse(LDO > LDO_mean + 2*LDO_sd | LDO < LDO_mean - 2*LDO_sd, "out", "in"),
         temp_filt = ifelse(Temp > Temp_mean + 2*Temp_sd | Temp < Temp_mean - 2*Temp_sd, "out", "in"))

```

```{r do-filter-all, echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, fig.height=5, fig.width=6, fig.cap="DO measurements from the sonde (points), with the ribbons showing the 2 standard deviations above and below the daily means, with the red points showing observations outside these bounds."}

# plot
sonde_raw %>%
  ggplot(aes(Date_Time, LDO))+
  facet_wrap(~year, scales = "free_x")+
  geom_ribbon(aes(ymin = LDO_mean - 2*LDO_sd, ymax = LDO_mean + 2*LDO_sd),
              alpha = 0.5, color = "gray50", linetype = 0)+
  geom_point(aes(color = do_filt), size = 0.3, alpha = 0.5)+
  scale_color_manual(values=c("gray20","red"),guide=FALSE)+
  ylab(expression(DO~"("*mg~O[2]~L^{-1}*")"))+
  xlab("Time")
```

```{r do-filter-subset, echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE,fig.height=5, fig.width=6, fig.cap="DO measurements for a sample of 20 days from the sonde (points), with the ribbons showing the 2 standard deviations above and below the daily means, with the red points showing observations outside these bounds."}

# plot
sonde_raw  %>%
  filter(year == 2017, !(yday < 175 | yday > 195)) %>%
  ggplot(aes(Date_Time, LDO))+
  facet_wrap(~year)+
  geom_ribbon(aes(ymin = LDO_mean - 2*LDO_sd, ymax = LDO_mean + 2*LDO_sd),
              alpha = 0.5, color = "gray50", linetype = 0)+
  geom_point(aes(color = do_filt), size = 0.5, alpha = 0.5)+
  scale_color_manual(values=c("gray20","red"),guide=FALSE)+
  ylab(expression(DO~"("*mg~O[2]~L^{-1}*")"))+
  xlab("Time")
```

\clearpage

## Sensor drift

```{r echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE}

# read clean sonde data
sonde = read_csv("data/sonde_final.csv") %>% 
  # convert hour from 0:23 to 1:24 (this makes indexing easier later)
  mutate(hour = hour + 1) 


# profiles
profile = read_csv("data/myvatn_profile.csv") %>%
  filter(sta %in% c(3, 33), sampledepth == 0.5) %>%
  mutate(year = year(sampledate),
         yday = yday(sampledate),
         hour = hour(sampletime)) %>%
  select(year, yday, hour, do) %>%
  rename(do_prof = do)

# combine sonde and profile data
comb_data = sonde %>%
              select(year, yday, hour, do) %>%
              rename(do_sonde = do) %>% na.omit() %>%
  left_join(profile)

# correlation
cor_do = cor.test(comb_data$do_prof, comb_data$do_sonde)
```

To evaluate whether the sonde DO readings drifted through time, I compared them to measurments and the same depth (0.5m) and location made with handheld probes (ProODO, YSI Co., OH, USA) during routine sampling of the central monitoring station. The sonde and handheld probe measurements were strongly correalted ($r$=0.92; Fig \ref{fig:sonde-vs-profile}). Furthermore, the DO time-series measured by the sonde failed to display linear trends that were consistent between years (Fig \ref{fig:sonde-profile-time}). Therefore, there did not appear to be any major drift in the sensor readings.

```{r sonde-vs-profile, echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, fig.height=5, fig.width=6, fig.cap="DO measured from the sonde and handheld probes, with the 1-to-1 line."}
comb_data %>%
  ggplot(aes(do_prof, do_sonde))+
  geom_point()+
  geom_abline(intercept = 0, slope = 1)+
  scale_y_continuous(expression(Sonde~DO~"("*mg~O[2]~L^{-1}*")"),
                     limits=c(9.5,13))+
  scale_x_continuous(expression(Handheld~Probe~DO~"("*mg~O[2]~L^{-1}*")"),
                     limits=c(9.5,13))+
  coord_equal()
```

```{r sonde-profile-time, echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, fig.height=5, fig.width=6, fig.cap="Hourly DO measurements plotted through time. The lines are the sonde measurements, while the points are the handheld probe measurements."}
comb_data %>%
  filter(!(year %in% c(2012,2014))) %>%
  mutate(time = yday + hour/24) %>%
  ggplot(aes(time, do_sonde))+
  facet_wrap(~year)+
  geom_line()+
  geom_point(aes(y = do_prof), color="red")+
  scale_y_continuous(expression(DO~"("*mg~O[2]~L^{-1}*")"))+
  scale_x_continuous("Day of Year")
```

\clearpage

## Depth variation in DO

To evaluate vertical variation in DO, in 2017 and 2018 we paired the sonde with a miniDOT (Precison Measurement and Engineering, Vista, CA, USA) dissolved oxygen and temperature logger positions 12cm above the sediment surface. The sonde and miniDOTs were calibrated against each other before deploying the sensors. There was some divergence between the sonde and the miniDOT, but overeall they were strongly correlated and therefore the sonde captured a majority of the temporal variation in DO throughout the water column.

```{r minidot, echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, fig.height=4, fig.width=6, fig.cap="DO measurements from sonde (0.5m below surface) and miniDOT (12cm above lake bottom)."}
read_csv("supplement/supp_data/mini_clean.csv") %>%
  mutate(yday = yday(datetime),
         hour = hour(datetime)) %>%
  filter(yday < 230, site == "st33") %>%
  mutate(hour = hour + 1) %>%
  select(site, year, yday, hour, do.pel, do.ben) %>%
  na.omit() %>%
  gather(layer, do, do.pel, do.ben) %>%
  mutate(layer = ifelse(layer == "do.pel", "Sonde (pelagic)", "miniDOT (benthic)")) %>%
  group_by(site, year, yday, layer) %>%
  summarize(do = mean(do, na.rm = T)) %>%
  ggplot(aes(yday, do, color = layer))+
  facet_wrap(~year)+
  geom_line(size = 0.8)+
  scale_color_manual("",values=c("firebrick","dodgerblue"))+
  scale_fill_manual("",values=c("firebrick","dodgerblue"))+
  scale_y_continuous(expression(DO~"("*mg~O[2]~L^{-1}*")"))+
  scale_x_continuous("Day of Year")+
  theme(legend.position = c(0.65, 0.85))
```

\clearpage

## Light attenuation

To estimate the light attenuation rate $c_{d,h}$, I regressed observe light attenuation calculated from weekly measurements of PAR at 0.5m depth intervals against turbidity measured simultaneously by the sonde. I then used this regression (R2 = 0.66; Fig \ref{fig:ext-turb}) to predict $c_{d,h}$ from turbidity for each time point in the sonde data.

```{r ext-turb, echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, fig.height=5, fig.width=6, fig.cap="Light attenuation estimated from 0.5m profiles plotted against turbidity measured from the sonde. The line shows the regression fit"}

# extract relevant data
# replace 0's with the lowest observed value (for upcoming log transformation)
profile_clean = read_csv("data/myvatn_profile.csv")  %>%
  filter(sta %in% c(3,33)) %>%
  mutate(year = year(sampledate), 
         yday = yday(sampledate),
         hour = hour(sampletime),
         light = ifelse(light>0,light,{read_csv("data/myvatn_profile.csv") %>% 
             filter(light >0)}$light %>% min))

# define function for calculating light extinction
ext_fun = function(x){
  if (nrow(x) < 2) {
    z <- NA
  } else {
    m = lm(log(light) ~ sampledepth, data = x)
    z1 = m$coef[['(Intercept)']]
    z2 = m$coef[['sampledepth']]
  }
  data_frame(sampledate = x$sampledate[1],
             year = x$year[1], 
             yday = x$yday[1], 
             hour = x$hour[1], 
             par0 = exp(z1),
             ext = -z2)
}

# calculate extinction
ext_data = profile_clean %>% 
  select(sampledate, year, yday, hour, sampledepth, light) %>%
  na.omit() %>%
  split(.$sampledate, .$hour) %>% 
  map_df(~ ext_fun(.x)) %>%
  # filter negative values of ext
  filter(ext > 0) %>%
  # add PAR (including for times missing in sonde_full)
  left_join(read_csv("supplement/supp_data/hobo_weather.csv") %>%
              select(year, yday, hour, par) %>%
              na.omit()) %>%
  # add daily averages of turbidity
  left_join(read_csv("supplement/supp_data/sonde_hourly.csv") %>%
              group_by(year, yday) %>%
              summarize(pcyv = mean(pcyv, na.rm=T),
                        turb = mean(turb, na.rm=T)))

# fit linear model to predict extinction from turbidity
ext_mod1 = lm(ext ~ turb, data = ext_data)

# generate predicted values
nd = ext_data %>% expand(turb = c(min(turb, na.rm=T), max(turb, na.rm=T)))
nd$ext = predict(ext_mod1, newdata = nd)

# plot water extinction coefficient vs turbidity
ext_data %>%
  na.omit() %>%
  ggplot(aes(turb, ext))+
  geom_point()+
  geom_line(data = nd)+
  scale_y_continuous(expression(Light~attenuation~(m^{-1})))+
  scale_x_continuous("Turbidity (volts)")
```

\clearpage

# Model fitting

## Scaling the model

### NEP

For fitting the model, I z-scored (centered on the mean and divided by the standard deviation) the observed DO concentrations and scaled the model accordingly. I then nondimensionalized the scaled model to remove the scale parameters (mixing depth and mean/standard deviation for the observed DO). This put the observed values and model parameters on unit scale, which improved the computaitonal efficiency and allowed for the use for weakly-informative priors on unit scale. I then back-transformed the scaled parameters to recover their values on the original scale.

The model for DO dynamics (eqn 9 in the main text) is:
$$
DO_{t+1} = DO_t+ \frac{\Delta h}{z_{mix}} \left(NEP_t+k_t(DO^{eq}_t-DO_t) \right)+\epsilon^{proc}_t
$$

where $DO_t$ is the DO concentration [$mg~O_2~m^{-3}$], $DO^{eq}_t$ is the DO concentration at saturation [$mg~O_2~m^{-3}$], $\Delta h$ = 1 hour, $NEP_t$ is the net ecosystem production [$mg~O_2~m^{-2}~h^{-1}$], $k$ [$m~h^{-1}$] is the rate of oxygen exchange, $z_{mix}$ is the mixing depth [$3.3m$], and $\epsilon_{proc}$ is the processes error [$mg~O_2~m^{-3}$]. For simplicity, only a single time index $t$ is used, rather than indexing by hour and day as in the main text.

For clarity, $DO$ is replaced with $y$ and $NEP$ with $\phi$, yielding

$$
y_{t+1} = y_t+ \frac{\Delta h}{z_{mix}}\left(\phi_t+k_t(y^{eq}_t-y_t) \right)+\epsilon^{proc}_t.
$$

Z-scoring the observed DO values by the mean ($\mu$) and standard deviation ($\tau$) of the full set of DO observations implies the following scaling of the model:

$$
\frac{y_{t+1}-\mu}{\tau} = \frac{1}{\tau} \left(y_t - \mu + \frac{\Delta h}{z_{mix}} \left(\phi_t+k_t(y^{eq}_t-y_t) \right)+\epsilon^{proc}_t \right)
$$

which can be expanded as

$$
= \frac{1}{\tau} y_t - \frac{1}{\tau}\mu + \frac{1}{\tau}\frac{\Delta h}{z_{mix}}\phi_t + \frac{1}{\tau} \frac{\Delta h}{z_{mix}} k_t (y^{eq}_t-y_t) + \frac{1}{\tau}\epsilon^{proc}_t. 
$$

Rearranging to express all "$y$" terms to resemble the lefthand side yields

$$
\frac{y_{t+1}-\mu}{\tau} = \frac{y_t-\mu}{\tau} + \frac{1}{\tau}\frac{\Delta h}{z_{mix}}\phi_t + \frac{\Delta h}{z_{mix}} k_t\left( \left(\frac{y^{eq}_t-\mu}{\tau} + \frac{\mu}{\tau} \right)- \left(\frac{y_t-\mu}{\tau} + \frac{\mu}{\tau}\right) \right) + \frac{1}{\tau}\epsilon^{proc}_t
$$

which simplifies to 

$$
\frac{y_{t+1}-\mu}{\tau} = \frac{y_t-\mu}{\tau} + \frac{1}{\tau}\frac{\Delta h}{z_{mix}}\phi_t + \frac{\Delta h}{z_{mix}} k_t\left( \frac{y^{eq}_t-\mu}{\tau} - \frac{y_t-\mu}{\tau} \right) + \frac{1}{\tau}\epsilon^{proc}_t. 
$$

Substituting with scaled variables yields

$$
x_{t+1} = x_t + \hat{\phi_t} + \hat{k}_t (x^{eq}_t - x_t) + \hat{\epsilon}^{proc}_t
$$
where 

\begin{equation}
\begin{split}
&x_t = \frac{y_t-\mu}{\tau}  \\
&x^{eq}_t = \frac{y^{eq}_t-\mu}{\tau} \\
&\hat{\phi}_t = \frac{1}{\tau}\frac{\Delta h}{z_{mix}}\phi_t \\
&\hat{k}_t = \frac{\Delta h}{z_{mix}} k_t \\
&\hat{\epsilon}^{proc}_t = \frac{1}{\tau}\epsilon^{proc}_t
\end{split}
\end{equation}

I fit the nondimensionalized version of the model, resulting in esitmates for the scaled parameters (e.g. $\hat{\phi}_t$). The parameter values on the original scale could be obtained by back-transforming the scaled values. For example, $\phi_t = \tau\frac{z_{mix}}{\Delta h} \hat{\phi}_t$.

###GPP and ER

NEP ($\phi_t$) is the difference between GPP and ER. Therefore, the scaling of NEP implies that GPP (denoted $\chi$) and ER (denoted $\kappa$) are scaled as 

\begin{equation}
\begin{split}
\hat{\chi}_t = \frac{1}{\tau}\frac{\Delta h}{z_{mix}}\chi_t\\
\hat{\kappa}_t = \frac{1}{\tau}\frac{\Delta h}{z_{mix}}\kappa_t\\
\end{split}
\end{equation}

The original equation for $\chi$ (main text eqn X) is

$$
\chi_t = \beta_{d}~\text{tanh} \left( \frac{\alpha_t}{\beta_t}L_t\right)  
$$ 

where $\beta_t$ is the maximum rate of GPP [$mg~O_2~m^{-3}~h^{-1}$] on a particular day and $\alpha$ [$mg~O_2~s~\mu mol~photons^{-1}~m^{-1}~h^{-1}$] is the initial slope of the P-I at low light. This is scaled as

$$
\hat{\chi}_t = \frac{1}{\tau}\frac{\Delta h}{z_{mix}}\beta_t~\text{tanh} \left( \frac{\frac{\lambda}{\tau}\frac{\Delta h}{z_{mix}}\alpha_t}{\frac{1}{\tau}\frac{\Delta h}{z_{mix}}\beta_t}\frac{L_t}{\lambda}\right)  
$$ 

where $\lambda$ is the mean light level (so that the initial slope can be expressed on unit scale). This simplifies to 

$$
\hat{\chi}_t = \hat{\beta}_t~\text{tanh} \left( \frac{\hat{\alpha}_t}{\hat{\beta}_t}\hat{L}_t\right)  
$$ 

where 
\begin{equation}
\begin{split}
&\hat{\beta}_t = \frac{1}{\tau}\frac{\Delta h}{z_{mix}}\beta_t\\
&\hat{\alpha}_t = \frac{\lambda}{\tau}\frac{\Delta h}{z_{mix}}~\alpha_t\\
&\hat{L}_t = \frac{L_t}{\lambda}.
\end{split}
\end{equation}

This scaling can be applied by direct analogy to ER and to the time varying components of $\beta_t$, $\alpha_t$, and $\rho_t$ (main text eqns X-X)

\clearpage

## Priors

I fit the model using weakly informative priors based on normal distributions (see http://mc-stan.org/users/documentation/case-studies/weakly_informative_shapes.html). Because the parameters were fit on unit scale, I generally used priors with standard deviations of 1 and means of 0. However, because the parameters describing the scaling of GPP and ER with temperature ($\gamma_\beta$ and $\gamma_\rho$) had lower bounds of 1, I used corresponding priors with means of 1. For those parameters with lower bounds (1 for $\gamma_\beta$ and $\gamma_\rho$ and 0 for all standard deviations), I used normal priors truncated at their respective lower bounds. 

\clearpage

## Estimating observation error

Attempting to estimate the standard deviaitons for observation error ($\sigma_{obs}$) and process error ($\sigma_{proc}$) independently resulted in very inefficient sampling of the posterior distribution as $\sigma_{obs}$ slowly converged to 0. This can be seen by examining the parameter trace plots from the MCMC (Fig \ref{fig:trace}). While $\sigma_{proc}$ showed good mixing across all four chains, the chains for $\sigma_{proc}$ did not mix, indicating a lack of convergence. However, the overall magnitude of $\sigma_{obs}$ was close to 0 and the lack of convergence is likely due to difficulties associated with estimating the parameter when near it's boundary.

Therefore, I refit the model fixing $\sigma_{obs}$ to near 0 (1% of the observed standard deviation in obseved DO, so that I could use the same model specification in Stan). This resulted in nearly identical estimates for most parameters except for $\sigma_{proc}$, which was nontheless very similar (Fig \ref{fig:posts-sig-obs}). It is not surprising that observation error was so low, given the smooth appearance of the DO time series (Fig \ref{fig:do}).

\clearpage

## Stan specifications and diagnostics

Detailed explanation of code for specifying, fitting, and diagnosing the model with Stan can be found at https://github.com/jsphillips2/sonde_oxygen.

```{r trace, echo = FALSE, message = FALSE, cache = FALSE, fig.height=4, fig.width=6, fig.cap="Trace plots for standard deviations of observation and process error from the MCMC. Colors indicate separately initialized chains (note the different y-axis scales)."}
read_csv("model/output/sig_obs/fixed_pars_full.csv") %>%
  select("chain","step","sig_obs","sig_proc") %>%
  rename("sigma[obs]" = "sig_obs", "sigma[proc]" = "sig_proc") %>%
  gather(par, value, -chain, -step) %>%
  ggplot(aes(step, value, color=factor(chain)))+
  facet_wrap(~par, scales="free_y", labeller = label_parsed)+
  geom_line(alpha=0.4)+
  scale_color_manual(values=c("blue","black","orange","red")) +
  ylab("Value")+
  xlab("Iteration")+
  guides(color = F)
```

```{r posts-sig-obs, echo = FALSE, message = FALSE, cache = FALSE, fig.height=6, fig.width=5, fig.cap="Posterior distributions for parameter estimates, with the standard deviation of observation error either estimated from data or fixed to 0."}
read_csv("model/output/sig_obs/fixed_pars_full.csv") %>%
  select(-"lp__",-"sig_obs") %>%
  mutate(type = "Estimated") %>%
  bind_rows(read_csv("model/output/main/fixed_pars_full.csv") %>%
  select(-"lp__") %>%
  mutate(type = "Fixed to 0")) %>%
  rename("gamma[beta]" = "gamma_1", "gamma[rho]" = "gamma_2",
         "sigma[alpha]" = "sig_a", "sigma[beta[0]]" = "sig_b0",
         "sigma[rho]" = "sig_r", "sigma[proc]" = "sig_proc") %>%
  gather(var, value, -step, -chain, -type) %>%
  ggplot(aes(value, linetype = type))+
  facet_wrap(~var, scales = "free", labeller = label_parsed)+
  stat_density(aes(linetype = type), position = "identity", geom = "line")+
  scale_y_continuous("Posterior Probability Density", breaks=NULL)+
  scale_x_continuous("Value", breaks=NULL)+
  scale_linetype_manual(expression(sigma[obs]), values=c(1,2))+
  theme(legend.position="top", legend.direction = "vertical")
```

```{r do, echo = FALSE, message = FALSE, cache = FALSE, fig.height=4, fig.width=5, fig.cap="Observed DO time-series for first six days in 2012."}
read_csv("model/input/main/sonde_prep.csv") %>%
  filter(year == 2013, yday %in% 165:170) %>%
  ggplot(aes(hour, do))+
  facet_wrap(~yday)+
  geom_line(na.rm=T)+
  geom_point(na.rm=T)+
  scale_y_continuous(expression(DO~"("*mg~O[2]~L^{-1}*")"))+
  scale_x_continuous("Hour", breaks = 8*c(0.5:3.5))
```

\clearpage

# Alternative parameterizations of k

The characterize exchange of oxygen between the water and atmosphere (main text equation 11), I used to alternative parameterizations of the piston velocity $k_{d,h}$, following Cole and Caraco (1998) and Cruisius and Wanninkhof (2003). Both approches use the same functional form (main text equation 12), but with different values for the constants. Specifically, Cole and Caraco (1998) used the values $k_{0} = 2.07$, $k_{1} = 0.215$, and $k_{2} = 1.7$, while Cruisius and Wanninkhof (2003) used $k_{0} = 0.168$, $k_{1} = 0.228$, and $k_{2} = 2.2$. 

While the alternative parameterizations resulted of $k_{d,h}$ resulted in very different estimates of the exchange of oxygen between the water and atmosphere, the estimates of NEP were very similar (Fig (Fig \ref{fig:flux-k}). Consequently, the parameters of the P-I curve also had very similar estimates between the two methods (Figs \ref{fig:beta0-rho-k} and \ref{fig:alpha-k}). 

```{r echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE}
# prepare data
flux_d <- lapply(list("main","alt_k"), 
                 function(x){read_csv(paste0("model/output/",x,"/summary_clean.csv")) %>%
                     filter(name %in% c("NEP","AIR")) %>%
                     left_join(read_csv(paste0("model/input/",x,"/sonde_prep.csv")) %>%
                                 filter(is.na(unique_day)==F) %>%
                                 group_by(year, yday) %>%
                                 summarize(day = unique(unique_day))) %>%
                     full_join(read_csv(paste0("model/input/",x,"/sonde_prep.csv")) %>%
                                 expand(year,yday,name=c("NEP","AIR"))) %>%
                     mutate(middle = middle/1000,
                            name = factor(name, levels=c("NEP","AIR"))) %>%
                     mutate(analysis = x)}) %>% 
  bind_rows() %>%
  mutate(analysis = ifelse(analysis=="alt_k","Crusius & Wanninkhof","Cole & Caraco"),
         analysis = factor(analysis, levels = c("Crusius & Wanninkhof","Cole & Caraco")),
         name = ifelse(name=="AIR","EXC","NEP"))

```

```{r flux-k, echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, fig.height=6, fig.width=6, fig.cap="Daily Net ecosystem production (NEP) and oxygen exchange with atmosphere (EXC) estimated using piston velocities from either Cole and Caraco (1998) and Cruisius and Wanninkhof (2003). Values are defined with respect to the water column (i.e negative values indicate a decrease in water column DO)."}
# prepare data
flux_d %>%
  ggplot(aes(yday, middle, color = analysis, linetype = name))+
  facet_wrap(~year)+
  geom_line(size = 0.5)+
  geom_hline(yintercept = 0, size = 0.3, alpha = 0.5)+
  scale_color_manual("",values=c("firebrick","dodgerblue"))+
  scale_linetype_discrete("")+
  scale_y_continuous(expression(Daily~DO~Flux~"("*g~O[2]~m^{-2}~d^{-1}*")"))+
  scale_x_continuous("Day of Year", breaks = c(170, 200, 230), limits = c(150, 250))+
  theme(legend.position = "top")
```

```{r beta0-rho-k, echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, fig.height=6, fig.width=6, fig.cap="Ecosystem metabolism rates estimated using piston velocities from either Cole and Caraco (1998) and Cruisius and Wanninkhof (2003)."}

# prepare data
beta_rho_d = read_csv("model/output/main/summary_clean.csv") %>%
  filter(name %in% c("beta0","rho")) %>%
  left_join(read_csv("model/input/main/sonde_prep.csv") %>%
              filter(is.na(unique_day)==F) %>%
              group_by(year, yday) %>%
              summarize(day = unique(unique_day))) %>%
  mutate(type = "Cole & Caraco") %>%
  bind_rows(read_csv("model/output/alt_k/summary_clean.csv") %>%
  filter(name %in% c("beta0","rho")) %>%
  left_join(read_csv("model/input/alt_k/sonde_prep.csv") %>%
              filter(is.na(unique_day)==F) %>%
              group_by(year, yday) %>%
              summarize(day = unique(unique_day))) %>%
    mutate(type = "Crusius & Wanninkhof"))

# plot
beta_rho_d %>%
  ggplot(aes(yday, middle, color=name))+
  facet_wrap(~year)+
  geom_line(aes(linetype = type, size = type), alpha = 0.7)+
  geom_text(data = tibble(year = 2015, yday = 187, 
                              name = c("beta0","rho"), 
                              middle = c(490, 240)), 
            aes(color = name), 
            label=c("Max GPP", "Basline ER"), 
            hjust = 0)+
    geom_text(data = tibble(year = 2015, yday = 187, 
                              name = c("beta0","rho"), 
                              middle = c(410,160)), 
            aes(color = name), 
            label=c("(beta^0)", "(rho)"), parse = T, 
            hjust = 0)+
  scale_y_continuous(expression(Metabolism~Parameter~"("*mg~O[2]~m^{-2}~h^{-1}*")"))+
  scale_color_manual("",values=c("dodgerblue","firebrick2"), guide = F)+
  scale_linetype_manual("",values=c(2,1))+
  scale_size_manual(values=c(0.7, 0.4))+
  scale_x_continuous("Day of Year", breaks = c(170, 200, 230), limits = c(150, 250))+
  guides(size = F)+
  theme(legend.position = "top")
```

```{r alpha-k, echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, fig.height=6, fig.width=6, fig.cap="Initial slope of the P-I curve estimated using piston velocities from either Cole and Caraco (1998) and Cruisius and Wanninkhof (2003)."}

# prepare data
alpha_d = read_csv("model/output/main/summary_clean.csv") %>%
  filter(name %in% c("alpha")) %>%
  left_join(read_csv("model/input/main/sonde_prep.csv") %>%
              filter(is.na(unique_day)==F) %>%
              group_by(year, yday) %>%
              summarize(day = unique(unique_day))) %>%
  mutate(type = "Cole & Caraco") %>%
  bind_rows(read_csv("model/output/alt_k/summary_clean.csv") %>%
  filter(name %in% c("alpha")) %>%
  left_join(read_csv("model/input/alt_k/sonde_prep.csv") %>%
              filter(is.na(unique_day)==F) %>%
              group_by(year, yday) %>%
              summarize(day = unique(unique_day))) %>%
    mutate(type = "Crusius & Wanninkhof"))

# plot
alpha_d %>%
  filter(name %in% c("alpha")) %>%
  ggplot(aes(yday, middle))+
  facet_wrap(~year)+
  geom_line(aes(linetype = type, size = type), alpha = 0.7)+
  scale_linetype_manual("",values=c(2,1))+
  scale_size_manual(values=c(0.7, 0.4))+
  scale_y_continuous(expression(Initial~Slope~(alpha)~"("*mg~O[2]~s~mu*mol-photons^{-1}~h^{-1}*")"))+
  scale_x_continuous("Day of Year", breaks = c(160, 190, 220))+
  guides(size = F)+
  theme(legend.position = "top")
```

\clearpage

# Surface vs. average water column PAR

Because primary production occurs throughout the water column and in the benthos, I fit the model using the average water column light, calcualted as 

$$
\begin{aligned}
&\overline{L_{d,h}(z)} = \frac{1}{z_\text{max}} \int_0^{z_\text{max}} L^0_{d,h}~e^{-c_{d,h}~z}~dz =  \frac{L^0_{d,h}(1-e^{-c_d,h~z_\text{max}})}{z_\text{max}~c_{d,h}}
\end{aligned}
$$
where $z$ [$m$] is the vertical position in the water column, $z_{max}$ is the water column depth, $L^0_{d,h}$ is the PAR at the water surface $(z=0)$, and $c_{d,h}$ is the light attenuation coefficient (we used $c$ rather than the standard $k_d$ to avoid notational confusion with gas exhange constants used below). For notational simplicity, we henceforth denote $\overline{L_{d,h}(z)}$ as $L_{d,h}$. To estimate $c_{d,h}$, I regressed observe light attenuation calculated from weekly measurements of PAR at 0.5m depth intervals against turbidity and used this regression (y = 0.43 + 0.059 $\times$ turbidity; df = 34; $R^2$ = 0.66) to predict $c_{d,h}$ from turbidity for each time point in the sonde data. 

The estimates $\overline{L_{d,h}(z)}$ have some uncertainty to do the inference of water clarity from turbidity. Furthermore, it is common in other studies of ecosystem metaboslim to use light as measured near the water surface. To evaluate the consequences of using $\overline{L_{d,h}(z)}$, I refit the model to surface light ($L^0_{d,h}$) and compared the estimates of the time-varying parameters to those for the original model fit. 

The estimates of baseline ER and maximum GPP were almost identical for both surface and water column light, as they should be given that they characterize metabolism when light is 0 (i.e. when surface and water column light are the same) and when light is saturating (i.e. when differences between surface and water column later are irrelevant) (Fig \ref{fig:beta0-rho}). The average value of the initial slope of the P-I curve was substantially lower for the model fit using surface light, which again was as expected since the surface light values were always higher than the water column values (Fig \ref{fig:alpha}). However, the temporal patterns were very similar, with the largest differences being relatively high values in 2015 and 2018 when using the water column light. These two years had substantial declines in water column light due to reduced water clarity associated with cyanobacterial blooms. 

```{r beta0-rho, echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, fig.height=6, fig.width=6, fig.cap="Ecosystem metabolism rates estimated using either surface or average water column light."}

# prepare data
beta_rho_d = read_csv("model/output/main/summary_clean.csv") %>%
  filter(name %in% c("beta0","rho")) %>%
  left_join(read_csv("model/input/main/sonde_prep.csv") %>%
              filter(is.na(unique_day)==F) %>%
              group_by(year, yday) %>%
              summarize(day = unique(unique_day))) %>%
  mutate(type = "Water Column") %>%
  bind_rows(read_csv("model/output/surface_par/summary_clean.csv") %>%
  filter(name %in% c("beta0","rho")) %>%
  left_join(read_csv("model/input/surface_par/sonde_prep.csv") %>%
              filter(is.na(unique_day)==F) %>%
              group_by(year, yday) %>%
              summarize(day = unique(unique_day))) %>%
    mutate(type = "Surface"))

# plot
beta_rho_d %>%
  ggplot(aes(yday, middle, color=name))+
  facet_wrap(~year)+
  geom_line(aes(linetype = type, size = type), alpha = 0.7)+
  geom_text(data = tibble(year = 2015, yday = 187, 
                              name = c("beta0","rho"), 
                              middle = c(490, 240)), 
            aes(color = name), 
            label=c("Max GPP", "Basline ER"), 
            hjust = 0)+
    geom_text(data = tibble(year = 2015, yday = 187, 
                              name = c("beta0","rho"), 
                              middle = c(410,160)), 
            aes(color = name), 
            label=c("(beta^0)", "(rho)"), parse = T, 
            hjust = 0)+
  scale_y_continuous(expression(Metabolism~Parameter~"("*mg~O[2]~m^{-2}~h^{-1}*")"))+
  scale_color_manual("",values=c("dodgerblue","firebrick2"), guide = F)+
  scale_linetype_manual("",values=c(2,1))+
  scale_size_manual(values=c(0.7, 0.4))+
  scale_x_continuous("Day of Year", breaks = c(170, 200, 230), limits = c(150, 250))+
  guides(size = F)+
  theme(legend.position = c(0.82,0.91))
```

```{r alpha, echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, fig.height=6, fig.width=6, fig.cap="Initial slope of the P-I curve estimated using either surface or average water column light."}

# prepare data
alpha_d = read_csv("model/output/main/summary_clean.csv") %>%
  filter(name %in% c("alpha")) %>%
  left_join(read_csv("model/input/main/sonde_prep.csv") %>%
              filter(is.na(unique_day)==F) %>%
              group_by(year, yday) %>%
              summarize(day = unique(unique_day))) %>%
  mutate(type = "Water Column") %>%
  bind_rows(read_csv("model/output/surface_par/summary_clean.csv") %>%
  filter(name %in% c("alpha")) %>%
  left_join(read_csv("model/input/surface_par/sonde_prep.csv") %>%
              filter(is.na(unique_day)==F) %>%
              group_by(year, yday) %>%
              summarize(day = unique(unique_day))) %>%
    mutate(type = "Surface"))

# plot
alpha_d %>%
  filter(name %in% c("alpha")) %>%
  ggplot(aes(yday, middle))+
  facet_wrap(~year)+
  geom_line(aes(linetype = type, size = type), alpha = 0.7)+
  scale_linetype_manual("",values=c(2,1))+
  scale_size_manual(values=c(0.7, 0.4))+
  scale_y_continuous(expression(Initial~Slope~(alpha)~"("*mg~O[2]~s~mu*mol-photons^{-1}~h^{-1}*")"))+
  scale_x_continuous("Day of Year", breaks = c(160, 190, 220))+
  guides(size = F)+
  theme(legend.position = c(0.14,0.91))
```

\clearpage

# Drivers of maximum GPP (Figure)

Maximum GPP at high light $\beta_d^0$ was quite variable, and it was the dominant contributor to variation in overall GPP and NEP. In the main text, I report results from linear models regressing $\beta_d^0$ against daily mean phycocyanin (as a measure of cyanobacterial abundance) and larval midge abundance. Phycocyanin data were available for all days for which there were estimates of $\beta_d^0$, while midge data were only available approximately weekly. Therefore, I fit two models: (1) including phycocyanin for all days, and (2) including both phycocyanin and midge abundance for those days when midge samples were collected. In the main text I show a figure relating $\beta_d^0$ against phycocyanin for all days. Here I show the corresponding figure for those days where midges were samples (Fig \ref{fig:midges}).

```{r midges, echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, fig.height=6, fig.width=6, fig.cap="Time-series of maximum GPP (as inferred from the model), phycocyanin, and midge larva."}

# prepare beta0 and phycocyanin data
beta0_phyc = read_csv("model/output/main/summary_clean.csv") %>%
  filter(name %in% c("beta0")) %>%
  left_join(read_csv("model/input/main/sonde_prep.csv") %>%
              filter(is.na(unique_day)==F) %>%
              filter(pcyv < 0.3) %>%
              group_by(year, yday) %>%
              summarize(day = unique(unique_day),
                        pcyv = mean(pcyv))) %>%
  select(year, yday, middle, pcyv) %>%
  rename(beta0 = middle) 

# prepare midge data
midges_summary = read_csv("data/midges.csv") %>%
              filter(sta %in% c(3, 33)) %>%
              mutate(year = year(sampledate),
                     yday = yday(sampledate)) %>%
              group_by(year, yday, coreid) %>%
              summarize(tanyt = sum(tanyt/fract_count),
                        chiro = sum(chiro/fract_count),
                        midges = tanyt + chiro) %>%
              group_by(year, yday) %>%
              summarize(tanyt = mean(tanyt, na.rm=T),
                            chiro = mean(chiro, na.rm=T),
                            midges = mean(midges, na.rm=T))

# combine 
beta0_phyc_midge = beta0_phyc %>%
  left_join(midges_summary) %>%
  na.omit()

# plot
beta0_phyc_midge %>%
  gather(var, val, beta0, pcyv, midges) %>%
  group_by(var) %>%
  mutate(val = (val - mean(val, na.rm=T))/sd(val, na.rm=T)) %>%
  ungroup() %>%
  mutate(var = factor(var, 
                      levels=c("beta0","pcyv","midges"), labels=c("Max GPP", "Phycocyanin", "Midges"))) %>%
  ggplot(aes(yday, val, color = var))+
  facet_wrap(~year)+
  geom_line()+
  scale_color_manual("",values = c("black","cyan4","magenta4"))+
  scale_y_continuous("Z-Score (dimensionless)")+
  scale_x_continuous("Day of Year", breaks = c(160, 190, 220))+
  theme(legend.position = c(0.14, 0.88))
```

\clearpage

# References

Cole, J. J., and N. F. Caraco. 1998. Atmospheric exchange of carbon dioxide in a low-wind oligotrophic lake measured by the addition of SF6. Limnology and Oceanography. 43: 647–656. doi:10.4319/lo.1998.43.4.0647

Crusius, J., and R. Wanninkhof. 2003. Gas transfer velocities measured at low wind speed over a lake. Limnology and Oceanography 48: 1010–1017. doi:10.4319/lo.2003.48.3.1010

Thimijan, R. W., and R. D. Heins. 1983. Photometric, Radiometric, and Quantum Light Units of Measure A Review of Procedures for Interconversion. HortScience 18: 5.











